\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-12 NAR_DRAFT.tex
More pass by Ryan.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
Somatic mutations accumulate spontaneously in all dividing tissues, and the local rate of mutation depends sensitively on the nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Tumor cells additionally acquire mutations through exposure to mutagens or degradation, via mutation or deletion - of cellular machinery responsible for the identification and repair of damage or replication errors. 
Many mutagens and DNA repair mechanism defects also have highly specific mutation signatures, such that they can be identified by observing the mutations in the tumor \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific processes driving mutations in a particular tumor generate a tumor's unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}
Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

Here we present an algorithm for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor as well as its mutation rate, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
We provide a simple implementation in R, \texttt{batcaver} that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.

\begin{figure}
  \begin{center}
  \includegraphics{figures/signatures_only.pdf}
  \end{center}
  \caption{Mutation profiles. 
  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
  }
  \label{NAR-sigfig}
 \end{figure}

\section{MATERIALS AND METHODS}
\subsection{Somatic variant calling base probability model}

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

\begin{equation}
  \label{eqn:1}
$$
  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
$$
\end{equation}
where the density $\textrm{f}_{m,\nu}(x_i)$ can take several forms.

Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e. $\mathrm{p}(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
\begin{equation}
  \label{eqn:2}
$$
  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
$$
\end{equation}
Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.

\subsection{Site-specific prior probability of mutation}
The probability which we have denoted $\mathrm{p}(m)$ in equation \ref{eqn:2} for compactness is more precisely the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $\mathrm{p}(m,M)$.
However, $\mathrm{p}(m,M)$ is not constant for every site in the genome, and follows a distribution conditional on the genomic context surrounding the site, denoted here $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' rule to further decompose as

\begin{equation}
  \label{eqn:full_prior}
  $$
  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
  $$
\end{equation}
We next show how to estimate the quantities in equation \ref{eqn:full_prior}.

\subsection{Estimation of the mutation profile}
There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context at a particular genomic site is made up of the identity of the reference base and the identity of the 3` and 5` flanking bases.
There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, resulting in $2 \cdot 16 = 32$ possible tri-nucleotide contexts.
We define a mutation context $C$ indexed by $c=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
For each context, a mutation can be to any of the three possible alternate alleles {\{A,(C/T),G\}}. 
We can thus define $\mathrm{S}$, the substitution type, indexed by $c=\{1 \dots 32\}$ contexts and $m = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
Using the substitution type $\mathrm{S}$, we can rewrite equation \ref{eqn:full_prior} as

\begin{equation}
  \label{eqn:detailed_prior}
  $$
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)},
  \end{array}
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S}_{m,c})$ is the probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

>>>>

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Fig. \ref{NAR-sigfig} \textbf{a-c}, shows mutation profiles from several tumors.
\rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
Our algorithm estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
Our implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
Our algorithm is simple, computationally inexpensive, and can be integrated into numerous other variant callers.
Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.

\begin{figure}
  \begin{center}
  \includegraphics{figures/signatures_only.pdf}

  \rngcomment{Need more basic labels for each panel. E.g. ``Real breast tumor'', ``Real accumulated myeloid leukemia'', ``Simulated concentrated profile''}
  \end{center}
  \caption{Tumor mutation profiles.
  \rngcomment{Add explanation of how to read a profile. ``In each panel, the histograms show...''}
  \textbf{(a)} The observed mutation profile of a breast tumor \cite{Shi2018}.
  \textbf{(b)} The observed mutation profile of an acute myeloid leukemia \cite{Griffith2015}.
  \textbf{(c)} \rngcomment{Add a third (and possibly fourth) real tumor, with a more concentrated profile. So we can illustrate that our simulations aren't out of line with reality. Ideally two tumors of same type that have noticeably different profiles.}
  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
  }
  \label{NAR-sigfig}
 \end{figure}

\section{MATERIALS AND METHODS}
\subsection{Somatic variant calling probability model}

At every site in the genome with non-zero coverage, Next Generation Sequencing produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is local read depth.
Variant callers use the data $\mathbf{x}$ to choose between competing hypotheses:
\begin{equation}
  \label{eqn:hypothesis}
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f}.
  \end{array}
\end{equation}
Here $m$ is any of the 3 possible alternate non-reference bases and $\nu$ is the variant allele frequency.
The maximum likelihood estimate of $\nu$ is simply $\hat{f}$, the number of variant reads divided by the local read depth.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given that hypothesis and the prior probability of that hypothesis. 
Assuming that reads are independent, this is
\begin{equation}\label{eqn:1}
  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
\end{equation}
where $\textrm{f}_{m,\nu}(x_i)$ is the probability model for reads.

Assuming that the identity of the alternate allele and its allele frequency are independent and that $\nu$ is uniformly distributed, Eq.~\ref{eqn:1} becomes
\begin{equation}  \label{eqn:2}
  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
\end{equation}
The focus of \batcave is to provide a tumor- and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.

\subsection{Site-specific prior probability of mutation}
The probability that we have denoted $\mathrm{p}(m)$ in Eq.~\ref{eqn:2} is more precisely the joint probability that a mutation has occurred $M$ and that it was to allele $m$, which we denote $\mathrm{p}(m,M)$.
But $\mathrm{p}(m,M)$ is not uniform across the genome.
Rather it depends on the local genomic context $C$, so its full form is $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' theorem to further decompose as 
\begin{equation}
  \label{eqn:full_prior}
  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
\end{equation}
We next show how to estimate the quantities in Eq.~\ref{eqn:full_prior}.

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the mutation profile (Fig.~\ref{NAR-sigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
